Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-25 RevisedSubmission/hines_hetland_2017_gji.tex
added more

% landing on the FOGM noise model

In this section we establish a noise model for the GNSS data. The
noise in GNSS data, specifically the temporally correlated noise, has
been thoroughly studied over the past two decades
\citep[e.g.,][]{Zhang1997, Mao1999, Williams2004, Langbein2008}. In
these studies, temporally correlated noise is described with some
combination of Brownian motion, a first-order Gauss-Markov (FOGM)
process, and/or flicker noise. There is some physical justification
for using Brownian motion as a noise model because it accurately
describes the power spectrum of motion resulting from instability in
geodetic monuments \citep[e.g.,][]{Wyatt1982, Wyatt1989}. In
particular, the power spectrums for Brownian motion and monument
motion are both proportional to $f^{-2}$, where $f$ is frequency.
Brownian motion (which is also known as a Weiner process), by
definition, contains a reference time at which the process begins.
Since there is no notion of when noise ``begins" in GNSS data, we do
not find Brownian motion to be an appropriate noise model. On the
other hand, a FOGM process also has a power spectrum that is
proportional to $f^{-2}$, above some corner frequency, and it has no
reference time (i.e., a FOGM process is stationary). In this study, we
describe the temporal covariance of GNSS noise with a FOGM process. We
do not explore spatial covariance in GNSS noise and simply assume that
it is spatially uncorrelated.

Here we
describe the time dependence of $\eta$ as a FOGM process and consider
$\eta$ to be spatially uncorrelated. A FOGM process is a solution to
the stochastic differential equation

\begin{equation}\label{eq:FOGMdef}
\dot{\eta}(t) + \alpha \eta(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The FOGM process
degenerates to the commonly used Brownian motion noise model under the
condition that $\alpha=0$ and $\eta(0) = 0$. Our spatially
uncorrelated noise model that satisfies eq. (\ref{eq:FOGMdef}) is a
Gaussian process with zero mean and the covariance function
\begin{equation}\label{eq:FOGM}
C_\eta\left((\pos,t),(\pos',t')\right) = 
\begin{cases}
\frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right), &\pos = \pos' \\
0, &\pos \neq \pos'
\end{cases}. 
\end{equation}

We constrain the hyperparameters for $\eta$, $\alpha$ and $\beta$,
with a set of 38 continuous GNSS stations in the Pacific Northwest
that are east of 121$^\circ$W.  These stations are sufficiently far
from the subduction zone that they are unlikely to contain transient
signal associated with SSEs.  We clean the data for these stations by
removing jumps at times of equipment changes, and we remove outliers
that have been detected with the algorithm described in Section
\ref{sec:Outlier}. We then find $\alpha$ and $\beta$ for each station
time series with the Restricted Maximum Likelihood (REML) method
\cite[e.g.,][]{Harville1974,Cressie1992,Hines2017}. The REML method
finds the hyperparameters, which we collectively refer to as
$\mitbf{\theta}$, that maximize the likelihood function
\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_*^T\mitbf{K}(\mitbf{\theta})\data_*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}.
\end{equation}
\citet{Harville1974} showed that choosing the hyperparameters which
maximize eq. (\ref{eq:REML}) is equivalent to choosing the
hyperparameters which maximize the probability of drawing $\data_*$
from $\data$. We use the REML method over the maximum likelihood
method \citep[e.g.,][]{Langbein1997} because the REML method accounts
for the improper prior that we assigned to $\mitbf{a}$
\citep{Hines2017}. We independently estimate $\mitbf{\theta}$ for each
station, and so $\data_*$ consists of displacements for an individual
station. We are assuming $u(p)=0$ when estimating the noise
hyperparameters for this section. The distribution of inferred
$\alpha$ and $\beta$ are shown in Figure \ref{fig:NoiseParams}. The
amplitude of FOGM noise, $\beta$, for the easting and northing
components is notably low and are clustered around 0.5 mm/yr$^{0.5}$.
The corresponding estimates of $\alpha$ tend to cluster around 0
yr$^{-1}$, suggesting that noise can be described with Brownian
motion. We also estimate hyperparameters for the vertical component of
displacements, under the hope that vertical deformation gradients
could reveal some geophysical signal. The amplitude of FOGM noise for
the vertical component is relatively large with a median value of 13.5
mm/yr$^{0.5}$.  The inferred values for $\alpha$ are also higher for
the vertical component with a median value of 8.21 yr$^{-1}$. In
Figure \ref{fig:NoiseSamples}, we use the median values of $\alpha$
and $\beta$ to generate two random samples of FOGM noise for each
component. The samples span five years and over these five years the
easting and northing samples drift by about 1 mm. In the context of
detecting SSEs, which produce several mm's of surface displacement on
the time-scale of weeks, the estimated FOGM noise for the easting and
northing component is negligible. In contrast, the estimated FOGM
noise for the vertical component is larger than the signal we would
expect from SSEs. We suspect that the higher amplitude for the FOGM
noise in the vertical component is accommodating for deficiencies in
our rather simple seasonal model. Based on this analysis, we
henceforth ignore temporally correlated noise in the easting and
northing component because of its low amplitude. We also do not use
vertical displacements because of the presumably low signal-to-noise
ratio.

@@

@@

We next determine appropriate hyperparameters for $X$ and each of the
three candidate covariance functions for $T$. First, we clean the GNSS
datasets by removing offsets at times of equipment changes and
removing outliers with the method describe in Section
\ref{sec:Outlier}. For the outlier detection algorithm, our prior
model, $u$, is chosen to have a length-scale and time-scale that is
able to approximately describe SSE displacements. We use the SE
covariance function for $X$ with length-scale $\ell = 100$ km, and we
use the Wendland covariance function for $T$, due to its computational
efficiency, with time-scale $\tau = 0.1$ yr and $\phi = 1$ mm.  The
outlier detection algorithm is particularly effective at removing
outliers for stations at high elevation (Figure \ref{fig:Outliers}),
which can be adversely affected by ice or snow during the winter
\citep{Lisowski2008}. After cleaning the dataset, we divide it into
seven subsets which are four months long and each centered on the time
of a SSE. The times of the seven SSEs are determined with tremor
records from \cite{Wech2010}. We use the REML method to find the
optimal hyperparameters for $T$ and $X$ for each subset of data. We
choose to make each data subsets four months long because it is long
enough to encompass a SSE in the Pacific Northwest, while it is short
enough to still be computationally tractable. However, four months is
too short to resolve the sinusoids in $\data$, and they are omitted
from $\data$ in this REML analysis for Pacific Northwest SSEs. The
estimated hyperparameters for $u$ are summarized in Table 1. Based on
the interquartile ranges, the estimated hyperparameters for the SE and
Wendland covariance functions do not vary significantly between SSEs.
This suggests that the median of estimated hyperparameters should be
an appropriate prior model for all Pacific Northwest SSEs. For the IBM
model, there are several anomalously large values for $\ell$ and
$\phi$, which results in large interquartile ranges.

\begin{figure*}
\includegraphics{figures/outliers/outliers.pdf}
\caption{
Detrended easting component of displacements at station SC03, which is
located on Mount Olympus in Washington. The orange markers indicate
outliers that were automatically detected using the algorithm from
Section \ref{sec:Outlier}. The error bars show one standard deviation
uncertainties. Note that outliers tend be observed in the winter,
suggesting that they were caused by snow or ice.
}   
\label{fig:Outliers}
\end{figure*}

@@

Having established a noise model and a prior for transient
displacements, we use the cleaned GNSS dataset to calculate transient
strain rates in the Puget Sound region.  We calculate transient strain
rates for each day from January 1, 2010 to May 15, 2017. The strain
rates are estimates at a grid of points spanning the study area. In
Figure \ref{fig:StrainMap} we show the transient strain rates on
January 1, 2016, which coincides with the height of the winter
2015-2016 SSE. We have included an animation showing the map view of
strain rates through time as supplementary material. The strain rates
shown in Figure \ref{fig:StrainMap} are generally similar to the
strain rates during the other six SSEs considered in this study. The
SSEs cause trench perpendicular compression in the Olympic Peninsula
and extension east of Puget Sound. The strain transitions from
compression to extension around the southern tip of Vancouver Island,
which coincides with the location of thrust slip for SSEs in the Puget
Sound region \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}. Thus,
this pattern of strain is to be expected. During the period in between
SSEs, secular strain rates indicate trench perpendicular compression
throughout this study region \citep{Murray2000, McCaffrey2007,
McCaffrey2013}. When comparing inferred strain rates from SSEs to the
secular strain rates, we see that SSEs are concentrating tectonically
accumulated strain energy towards the trench, and presumably pushing
the subduction zone closer to failure.

>>>>

% Defining the components of the noise vector

We consider the noise vector, $\mitbf{\eta}_i$, to be composed of a
temporally correlated Gaussian process, $\mathcal{L}_i \sim
\mathcal{GP}(0,C_{\mathcal{L}_i})$, and a vector of uncorrelated
Gaussian noise, $\mitbf{w}_i$, so that
\begin{equation}
\mitbf{\eta}_i = \mathcal{L}_i(\points) + \mitbf{w}_i.
\end{equation}
The standard deviations for $\mitbf{w}_i$ are given by the
uncertainties derived for the GNSS displacement solutions,
$\mitbf{\sigma}_i$. The noise vector then has zero mean and the
covariance matrix
\begin{equation}
\mitbf{C}_{\eta_i} = C_{\mathcal{L}_i}(\points,\points) +
                     \mathrm{diag}(\mitbf{\sigma}_i^2).
\end{equation}

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades \citep[e.g.,][]{Zhang1997, Mao1999,
Williams2004, Langbein2008}. In these studies, temporally correlated
noise tends to be described with some combination of Brownian motion
(also known as random walk noise or a Weiner process), a first-order
Gauss-Markov (FOGM) process, and/or flicker noise. There is some
physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. In particular, the power spectrums for Brownian motion and
monument motion are both proportional to $f^{-2}$, where $f$ is
frequency. Brownian motion, by definition, contains a reference time
at which the process begins. Since there is no notion of when noise
``begins" in GNSS data, we do not find Brownian motion to be an
appropriate noise model. On the other hand, a FOGM process also has a
power spectrum that is proportional to $f^{-2}$, above some corner
frequency, and it has no reference time (i.e., a FOGM process is
stationary) \citep{Langbein2004}. We then let $\mathcal{L}_i$ be a
spatially uncorrelated FOGM process, which has the covariance function
\begin{equation}\label{eq:FOGM}
C_{\mathcal{L}_i}\left((\pos,t),(\pos',t')\right) = 
\frac{\beta^2}{2\alpha}
\exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'}
\end{equation}
\citep[sec. B2.1]{Rasmussen2006}, where $\delta_{\pos,\pos'}$ is 1 if
$\pos = \pos'$ and 0 otherwise.

% How we constrain the parameters

We constrain the parameters for $\mitbf{\eta}_i$, $\alpha$ and
$\beta$, with a set of 38 continuous GNSS stations in the Pacific
Northwest that are east of 121$^\circ$W.  These stations are
sufficiently far from the subduction zone that they are unlikely to
contain transient signal associated with SSEs. We find $\alpha$ and
$\beta$ for each station time series with the Restricted Maximum
Likelihood (REML) method \cite[e.g.,][]{Harville1974, Cressie1992}.
The REML method finds the hyperparameters, which we collectively refer
to as $\mitbf{\theta}$, that maximize the likelihood function
\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_*^T\mitbf{K}(\mitbf{\theta})\data_*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}.
\end{equation}
\citet{Harville1974} showed that choosing the hyperparameters which
maximize eq. (\ref{eq:REML}) is equivalent to choosing the
hyperparameters which maximize the probability of drawing $\data_*$
from $\data$. We use the REML method over the maximum likelihood
method \citep[e.g.,][]{Langbein1997} because the REML method accounts
for the improper prior that we assigned to $\mitbf{a}$
\citep{Hines2017}. We independently estimate $\mitbf{\theta}$ for each
station, and so $\data_*$ consists of displacements for an individual
station. We are assuming $u(p)=0$ when estimating the noise
hyperparameters for this section.

% results of estimated parameters

The distribution of inferred $\alpha$ and $\beta$ are shown in Figure
\ref{fig:NoiseParams}. The amplitude of FOGM noise, $\beta$, for the
easting and northing components is notably low and are clustered
around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend
to cluster around 0 yr$^{-1}$, suggesting that noise can be described
with Brownian motion. We also estimate hyperparameters for the
vertical component of displacements, under the hope that vertical
deformation gradients could reveal some geophysical signal. The
amplitude of FOGM noise for the vertical component is relatively large
with a median value of 13.5 mm/yr$^{0.5}$.  The inferred values for
$\alpha$ are also higher for the vertical component with a median
value of 8.21 yr$^{-1}$. In Figure \ref{fig:NoiseSamples}, we use the
median values of $\alpha$ and $\beta$ to generate two random samples
of FOGM noise for each component. The samples span five years and over
these five years the easting and northing samples drift by about 1 mm.
In the context of detecting SSEs, which produce several mm's of
surface displacement on the time-scale of weeks, the estimated FOGM
noise for the easting and northing component is negligible. In
contrast, the estimated FOGM noise for the vertical component is
larger than the signal we would expect from SSEs. We suspect that the
higher amplitude for the FOGM noise in the vertical component is
accommodating for deficiencies in our rather simple seasonal model.
Based on this analysis, we henceforth ignore temporally correlated
noise in the easting and northing component because of its low
amplitude. We also do not use vertical displacements because of the
presumably low signal-to-noise ratio.

@@

% Note spatially correlated noise

@@

We next determine appropriate hyperparameters for $X$ and each of the
three candidate covariance functions for $T$. we divide the dataset
into seven subsets which are four months long and each centered on the
time of a SSE. The times of the seven SSEs are determined with tremor
records from \cite{Wech2010}. We use the REML method to find the
optimal hyperparameters for $T$ and $X$ for each subset of data. We
choose to make each data subsets four months long because it is long
enough to encompass a SSE in the Pacific Northwest, while it is short
enough to still be computationally tractable. However, four months is
too short to resolve the sinusoids in $\data$, and they are omitted
from $\data$ in this REML analysis for Pacific Northwest SSEs. The
estimated hyperparameters for $u$ are summarized in Table 1. Based on
the interquartile ranges, the estimated hyperparameters for the SE and
Wendland covariance functions do not vary significantly between SSEs.
This suggests that the median of estimated hyperparameters should be
an appropriate prior model for all Pacific Northwest SSEs. For the IBM
model, there are several anomalously large values for $\ell$ and
$\phi$, which results in large interquartile ranges.

@@

Having established a noise model and a prior for transient
displacements, we use the GNSS dataset to calculate transient strain
rates in the Puget Sound region. We calculate transient strain rates
for each day from January 1, 2010 to May 15, 2017. The strain rates
are estimates at a grid of points spanning the study area. In Figure
\ref{fig:StrainMap} we show the transient strain rates on January 1,
2016, which coincides with the height of the winter 2015-2016 SSE. We
have included an animation showing the map view of strain rates
through time as supplementary material. The strain rates shown in
Figure \ref{fig:StrainMap} are generally similar to the strain rates
during the other six SSEs considered in this study. The SSEs cause
trench perpendicular compression in the Olympic Peninsula and
extension east of Puget Sound. The strain transitions from compression
to extension around the southern tip of Vancouver Island, which
coincides with the location of thrust slip for SSEs in the Puget Sound
region \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}. Thus, this
pattern of strain is to be expected. During the period in between
SSEs, secular strain rates indicate trench perpendicular compression
throughout this study region \citep{Murray2000, McCaffrey2007,
McCaffrey2013}. When comparing inferred strain rates from SSEs to the
secular strain rates, we see that SSEs are concentrating tectonically
accumulated strain energy towards the trench, and presumably pushing
the subduction zone closer to failure.
